---
title: "Derive New Variables with dplyr"
tutorial:
  id: "io.shinyapps.tutorials.03a-deriving"
  version: 0.01
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    df_print: default
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(babynames)

checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
knitr::opts_chunk$set(echo = FALSE)
```

```{r bnames, include=FALSE, exercise=TRUE}
```


## Welcome

In this case study, you will identify the most popular American names from 1880 to 2015. While doing this, you will master three more dplyr functions:

* `summarize()`, which help you use your data to compute summary statistics (akin to Stata's `collapse`), 
* `mutate()`, which help you to append new variables to your dataset as a combination of existing ones (akin to Stata's `generate`). 

The key difference between these two functions and the previous dplyr verbs is that they **create new columns**  in your dataset instead of extracting existing information. 


Note: This tutorial is a modification of the original RStudio Primers, which contained more information about `summarize()` and `group_by()` that was not as crucial for a preparatory tutorial. It uses the [core tidyverse packages](http://tidyverse.org/), including ggplot2, tibble, and dplyr, as well as the `babynames` package. All of these packages have been pre-installed and pre-loaded for your convenience.

Click the Next Topic button to begin.

## The most popular names

### What are the most popular names of all time?

Let's use `babynames` to answer a different question: what are the most popular names of all time?

This question seems simple enough, but to answer it we need to be more precise: how do you define "the most popular" names? Try to think of several definitions and then click Continue. After the Continue button, I will suggest two definitions of my own.

### Two definitions of popular

I suggest that we focus on two definitions of _popular_, one that uses sums and one that uses ranks:

1. **Sums** - A name is popular _if the total number of children that have the name is large when you sum across years_.
2. **Ranks** - A name is popular _if it consistently ranks among the top names from year to year_.

This raises a question:

```{r q1}
question(
  "Do we have enough information in `babynames` to compare the popularity of names?",
  answer("No. No cell in `babynames` contains a rank value or a sum across years."),
  answer("Yes. We can use the information in `babynames` to compute the values we want.", correct = TRUE),
  allow_retry = TRUE
)

```

### Deriving information

Every data frame that you meet implies more information than it displays. For example, `babynames` does not display the total number of children who had your name, but `babynames` certainly implies what that number is. To discover the number, you only need to do a calculation:

```{r echo = TRUE}
babynames |> 
  filter(name == "Garrett", sex == "M") |> 
  summarize(total = sum(n))
```

### Useful functions

dplyr provides three functions that can help you reveal the information implied by your data: 

* `summarize()`  
* `group_by()`
* `mutate()`

Like `select()`, `filter()` and `arrange()`, these functions all take a data frame as their first argument and return a new data frame as their output, which makes them easy to use in pipes.

Let's master each function and use them to analyze popularity as we go.

## summarize()

`summarize()` takes a data frame and uses it to calculate a new data frame of summary statistics. 

### Syntax

To use `summarize()`, pass it a data frame and then one or more named arguments. Each named argument should be set to an R expression that generates a single value. summarize will turn each named argument into a column in the new data frame. The name of each argument will become the column name, and the value returned by the argument will become the column contents.

### Example

I used `summarize()` above to calculate the total number of boys named "Garrett", but let's expand that code to also calculate 

* `max` - the maximum number of boys named "Garrett" in a single year
* `mean` - the mean number of boys named "Garrett" per year

```{r echo = TRUE}
babynames |> 
  filter(name == "Garrett", sex == "M") |> 
  summarize(total = sum(n), max = max(n), mean = mean(n))
```

Don't let the code above fool you. The first argument of `summarise()` is always a data frame, but when you use `summarise()` in a pipe, the first argument is provided by the pipe operator, `|>`. Here the first argument will be the data frame that is returned by `babynames |> filter(name == "Garrett", sex == "M")`.

### Exercise - summarise()

Use the code chunk below to compute three statistics:

1. the total number of children who ever had your name
1. the maximum number of children given your name in a single year
1. the mean number of children given your name per year 

If you cannot think of an R function that would compute each statistic, click the Hint/Solution button.

```{r summarise-1, exercise = TRUE, exercise.setup = "bnames"}

```

```{r summarise-1-solution}
babynames |> 
  filter(name == "Garrett", sex == "M") |> 
  summarise(total = sum(n), max = max(n), mean = mean(n))
```

### Summary functions

So far our `summarise()` examples have relied on `sum()`, `max()`, and `mean()`. But you can use any function in `summarise()` so long as it meets one criteria: the function must take a _vector_ of values as input and return a _single_ value as output. Functions that do this are known as **summary functions** and they are common in the field of descriptive statistics. Some of the most useful summary functions include:

1. **Measures of location** - `mean(x)`, `median(x)`, `quantile(x, 0.25)`, `min(x)`, and `max(x)`
1. **Counts** - `n_distinct(x)` and `n()`, which takes no arguments, and returns the size of the current group or data frame. 
1. **Counts and proportions of logical values** - `sum(!is.na(x))`, which counts the number of `TRUE`s returned by a logical test; `mean(y == 0)`, which returns the proportion of `TRUE`s returned by a logical test.

Let's apply some of these summary functions. Click Continue to test your understanding. 

### Khaleesi challenge

"Khaleesi" is a very modern name that appears to be based on the _Game of Thrones_ TV series, which premiered on April 17, 2011. In the chunk below, filter babynames to just the rows where name == "Khaleesi". Then use `summarise()` and a summary function to return the first value of `year` in the data set.

```{r khaleesi, exercise = TRUE, exercise.setup = "bnames"}

```

```{r khaleesi-solution}
babynames |> 
  filter(name == "Khaleesi") |> 
  summarise(year = first(year))
```

### Distinct name challenge

In the chunk below, use `summarise()` and a summary function to return a data frame with two columns:

* A column named `n` that displays the total number of rows in `babynames`
* A column named `distinct` that displays the number of distinct names in `babynames`

Will these numbers be different? Why or why not?

```{r distinct, exercise = TRUE, exercise.setup = "bnames"}

```

```{r distinct-solution}
babynames |> 
  summarise(n = n(), distinct = n_distinct(name))
```

```{r distinct-check}
"Good job! The two numbers are different because most names appear in the data set more than once. They appear once for each year in which they were used."
```


## mutate()

The `summarize()` command provides summaries of existing variables, but we often want to transform our variables without summarizing them. These can be as simple as translating the units of variables from meters to kilometers, or creating a new variable that is the sum or ratio of the other two variables. 

`mutate()` uses a data frame to compute new variables. It then returns a copy of the data frame that includes the new variables. For example, we can use `mutate()` to compute a `percent` variable for `babynames`. Here percent is just the `prop` multiplied by 100 and rounded to two decimal places.

```{r echo = TRUE}
babynames |>
  mutate(percent = round(prop * 100, digits =  2))
```

### Exercise - mutate()

The syntax of mutate is similar to `summarise()`. `mutate()` takes first a data frame, and then one or more named arguments that are set equal to R expressions. `mutate()` turns each named argument into a column. The name of the argument becomes the column name and the result of the R expression becomes the column contents.

Use `mutate()` in the chunk below to create a `births` column, the result of dividing `n` by `prop`. You can think of `births` as a sanity check; it uses each row to double check the number of boys or girls that were born each year. If all is well, the numbers will agree across rows (allowing for rounding errors).

```{r mutate-1, exercise = TRUE, exercise.setup = "bnames"}

```

```{r mutate-1-solution}
babynames |> 
  mutate(births = n / prop)
```

## Vectorized functions

Like `summarise()`, `mutate()` works in combination with a specific type of function. `summarise()` expects summary functions, which take a vector of input and return a _single_ value. `mutate()` expects **vectorized functions**, which take a vector of input and return a _vector_ of values.

In other words, summary functions like `min()` and `max()` won't work well with `mutate()`. You can see why if you take a moment to think about what `mutate()` does: `mutate()` adds a new column to the original data set. In R, every column in a dataset must be the same length, so `mutate()` must supply as many values for the new column as there are in the existing columns. 

If you give `mutate()` an expression that returns a single value, it will follow R's recycling rules and repeat that value as many times as needed to fill the column. This can make sense in some cases, but the reverse is never true: you cannot give `summarise()` a vectorized function; `summarise()` needs its input to return a single value.

What are some of R's vectorized functions? Click Continue to find out.

### The most useful vectorized functions

Some of the most useful vectorised functions in R to use with `mutate()` include:

1. **Arithmetic operators** - `+`, `-`, `*`, `/`, `^`. These are all vectorised, using R's so called "recycling rules". If one vector of input is shorter than the other, it will automatically be repeated multiple times to create a vector of the same length. 
1. **Modular arithmetic**: `%/%` (integer division) and `%%` (remainder)
1. **Logical comparisons**, `<`, `<=`, `>`, `>=`, `!=`
1. **Logs** - `log(x)`, `log2(x)`, `log10(x)`

## Recap

In this primer, you learned three functions for isolating data within a table: 

* `select()`
* `filter()`
* `arrange()`

You also learned three functions for deriving new data from a table: 

* `summarise()`
* `mutate()`

Together these six functions create a _grammar_ of data manipulation, a system of verbs that you can use to manipulate data in a sophisticated, step-by-step way. These verbs target the everyday tasks of data analysis. No matter which types of data you work with, you will discover that:

1. _Data sets often contain more information than you need_
1. _Data sets imply more information than they display_

The six dplyr functions help you work with these realities by isolating and revealing the information contained in your data. In fact, dplyr provides more than six functions for this grammar: dplyr comes with several functions that are variations on the themes of `select()`, `filter()`, `summarize()`, and `mutate()`. Each follows the same pipeable syntax that is used throughout dplyr. If you are interested, you can learn more about these peripheral functions in the [dplyr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf).



### Where to from here

Congratulations! You can use dplyr's grammar of data manipulation to access any data associated with a table---even if that data is not currently displayed by the table.

In other words, you now know how to look at data in R, as well as how to access specific values, calculate summary statistics, and compute new variables. When you combine this with the visualization skills that you learned in [Visualization Basics](http://tutorials.shinyapps.io/02-Vis-Basics), you have everything that you need to begin exploring data in R.

The next tutorial will teach you the last of three basic skills for working with R:

1. How to visualize data
1. How to work with data
1. How to program with R code





